Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-06-22 doc/paper.Rmd
intro smooth out, added a new citation

Tropical deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015; @alroy2017]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with forest remnants of different sizes embedded in a matrix of often highly-contrasting habitat [@taubert2018; @bianchi2013]. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008]. Fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000]. Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] are driving declines in reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998]. Despite the fact that this mechanism is thought to be particularly important in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions remains scarce.

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], with effects that are more pronounced on fragmented populations [@oliver2015; @holyoak2016]. While the demographic consequences are expected to be similarly severe in the tropics [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for species found in tropical biomes [@paniw2021]. Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015]. Plant populations may be particularly sensitive---an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019]. This increased sensitivity to climate fluctuations, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

@@

Climate models for the Amazon predict a future of extremes---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015]. Our results support the hypothesis that populations in fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014]. However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated. Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability. Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms. This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.

>>>>

The expansion of agriculture and other human activities is a primary driver of deforestation in the tropics [@haddadHabitatFragmentationIts2015; @alroy2017]. It also results in landscapes where the remaining forest can be highly fragmented, with patches of different sizes embedded in a matrix of often contrasting habitat [@taubert2018; @bianchi2013]. This fragmentation is associated with myriad ecological changes, including the local and regional extinction of plant species [@laurance2006; @dasilva2000]. Although the demographic mechanisms responsible these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009], it is often hypothesized that the dramatically altered environmental conditions in tropical forest fragments [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] drive declines in plant reproduction, recruitment, or survivorship [@brunaSeedGerminationRainforest1999; @zartmanDroughtImpactsSurvivorship2015; @lauranceRainForestFragmentation1998]. Despite the prevalence of this hypothesis [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999; @bettsExtinctionFiltersMediate2019], efforts to link population-level demographic responses with altered environmental conditions in fragments remains scarce.

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwoodEffectsClimateChange2015; @williams2015], and that the effects of climate change could be more pronounced when habitat is fragmented [@oliver2015; @holyoak2016]. While the demographic consequences of climate change for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the responses of these species to climatic variability [@paniw2021]. Tropical plants may be particularly sensitive to climate change---they typically have narrow ranges of climatic tolerance [@feeley2012], and recent results suggest increases in the frequency and severity of extreme precipitation events reduce survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019]. This sensitivity to climatic fluctuations, coupled with evidence that plant growth and survivorship are lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in forest fragments will be especially susceptible to climate change [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015].

@@

Over 24 million ha of the Brazilian Amazon have been cleared in the last two decades [@silvajunior2021], resulting in their extensive fragmentation [@broadbentForestFragmentationEdge2008]. Climate models predict a future of extremes for these forests---increases in the frequency and geographic extent of droughts, but also increases in the frequency and area affected by periods of unusual wetness [@duffy2015]. Our results support the hypothesis that populations in Amazonian forest fragments could be more susceptible to the effects of changing climate than those in continuous forest [@lauranceApparentEnvironmentalSynergism2014]. However, they also indicate that the demographic responses to climate change of populations in fragmented landscapes may be far more complex than previously appreciated. Multi-factorial, multi-season experiments [*sensu* @brunaCompensatoryResponsesUnderstory2005; @markewitz2010; @westerbandEarlyLifeConditions2017; @aguirre2021], ideally manipulating multiple combinations of climatic variables [@mundim2016], are needed to determine how and why habitat-specific differences in environmental conditions interact to delay the demographic responses of plants to climatic variability. Also needed are statistical tools that can test for synergistic effects of fragmentation and climate in vital rates, as those currently available do not allow for including interaction terms. This also limits the ability to include size by climate interactions in a DLNM; although plant responses to both fragmentation and climatic extremes can be size-specific [@brunaDemographicEffectsHabitat2005; @schwartzFragmentationForestStructure2019].
The ability to identify size-specific lagged responses may be especially complicated given size and growth are rarely measured at the same time scale as SPEI and other putative climatic drivers.
